Chain separation distribution in protein contact networks

### 2020-08-04 paper/main.tex
Updates from Overleaf

%\documentclass{article}
%\usepackage[a4paper, total={6in, 8in}]{geometry}
\documentclass[reprint,amsmath,amssymb,rmp,onecolumn,notitlepage,11pt]{revtex4-1}
\usepackage[utf8]{inputenc}
%\usepackage{authblk}
%\usepackage{natbib}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{xcolor}
\newcommand{\red}[1]{\textcolor{red!80!black}{#1}}
\newcommand{\blue}[1]{\textcolor{blue!80!black}{#1}}
\newcommand{\green}[1]{\textcolor{green!70!black}{#1}}
\newcommand{\gray}[1]{\textcolor{gray!80!black}{#1}}
\usepackage{mathtools}
\DeclarePairedDelimiter{\evdel}{\langle}{\rangle}
\usepackage[colorinlistoftodos]{todonotes}

\begin{document}
\title{The role of geometry in the chain separation distribution of links in protein contact networks}
\title{\red{We need a better title, the old one is too long} \green{Yes! But no better ideas at the moment}}
\title{\red{Chain separation distribution in protein contact networks}}
%\title{}
\author{Nora Molkenthin}
\affiliation{Potsdam Institut für Klimafolgenforschung}
\affiliation{Center for Advancing Electronics Dresden (cfaed), Technical University of Dresden}
\author{Steffen Mühle}
\affiliation{Uni Göttingen, 37077 Göttingen, Germany}
\author{Antonia S J S Mey}
\email[Electronic Address: ]{antonia.mey@ed.ac.uk}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}
%\author{Marc Timme}
% \affiliation{Network Dynamics, Max Planck Institute for Dynamics and Self-Organization (MPIDS), 37077 Göttingen, Germany}
% \affiliation{Chair for Network Dynamics, Institute for Theoretical Physics and Center for Advancing Electronics Dresden (cfaed), Technical University of Dresden, 01069 Dresden}

@@

Based on such considerations, the network formation process, repeatedly \red{(this may not be the adverb i'm looking for, but subsequently also doesnt quite work)} connects random pairs of units, without causing any intersections along the chain ~\cite{molkenthin2016scaling}. Such a model can now be used to investigate the emergent behaviour of sequence separation distributions as observed by Bartoli et al.~\cite{bartoli2008effect}. To this end, we analyze the sequence separation distributions that emerge from the geometric model, where the \emph{sequence separation} $\ell$ of a link is defined as the distance of the two end units along the original chain.
This is then compared to the sequence separation distribution extracted from protein residue networks (PRN) generated from protein data bank data.

@@

\section*{Protein Residue Networks and their synthetic counterparts}
The complex interaction pattern between the amino acids in a protein can be very naturally expressed as a network, in which each amino acid is represented by a node and spatial proximity is encoded as a link. Whenever two central $C_\alpha$ atoms are closer together than a threshold $d_c$, they are connected. These connections are encoded in the adjacency matrix of the folded protein, which is the binary matrix
\begin{equation}
  A^{\textsf{PDB}}_{ij}=
  \begin{cases}
   0, & \text{ if } d_{i,j}>d_c \text{ or } i=j\\
      1, & \text{ if } d_{i,j}\leq d_c .
      \end{cases}
    \label{eq:aij}
\end{equation}
Adjacency matrices are commonly used to describe networks and in this case represent the PRN.

The underlying formation mechanisms of individual protein folds are incredibly complex and depend on the surrounding solvent as well as the specific amino acid sequence and their quantum-mechanical interactions. Here we want to study the ensemble of all proteins, however, and will thus "average out" such specifics.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.4\columnwidth]{figures/Fig1.pdf}
    \caption{a) The sequence separation distribution is defined as the separation along the chain (blue) of two connected units (red). b) The solid blue link prohibits some connections from node $A$ (dotted red), while others are still permitted (dotted blue). Counting all newly excluded links, we find that a link of length $k$ prohibits $2(l-1)$ links of length $l$ up to a maximum of $k$ and $2(k-1)$ links of length $k$ or more.}
    \label{fig:schematic}
\end{figure}

@@

In subsequent steps we get the same reduction but starting from the pool $F_{k-1}(s)$ of the step before rather than the full $N$ links for each length. The reduction also needs to be multiplied by a factor 
\begin{equation}
    P_k(s)=\frac{F_k(s)}{C_k},
    \label{eq.Pk}
\end{equation}
with $C_k=\sum_{s=2}^{N/2}F^k(s)$, to only remove links that are still in the pool. This leads to the recursion
\begin{equation}
    F_k(s)= \frac{1}{N/2-2} \sum_{s_k=2}^{N/2} {\begin{cases}
     F_{k-1}(s)-2(s-1) P_{k-1}(s) \text{ , for } s<s_k\\
     F_{k-1}(s)-2(s_k -1)P_{k-1}(s)\text{ , for } s\geq s_k
    \end{cases}},
\end{equation}
which can be simplified to
\begin{align}
   F_k(s)&= \frac{1}{\frac{N}{2}-2} \frac{F_{k-1}(s)}{C_{k-1}}\left( \sum_{s_k=2}^{N/2}C_{k-1} - \sum_{s_k=2}^{s} 2(s_k-1) - \sum_{s_k=s+1}^{N/2} 2(s -1) \right) \nonumber \\
   &= \frac{1}{\frac{N}{2}-2}\frac{F_{k-1}(s)}{C_{k-1}}\left(\left(\frac{N}{2}-2\right)C_{k-1} -(s^2-s)-2\left(\frac{N}{2}-(s+1)\right)(s-1)\right)  \nonumber \\
   &= F_{k-1}(s)\left(1-\frac{-s^2 +(N-1)s-(N+1)}{(\frac{N}{2}-2)C_{k-1}} \right)\nonumber \\
   &=F_{k-1}(s)\left(1-\frac{f(s)}{(\frac{N}{2}-2)C_{k-1}} \right)
   \label{eq.Fk_rec}
\end{align}
Where we defined $f(s)=-s^2 +(N-1)s-(N+1) \approx N s - s^2 - N$ for notational brevity.

@@

In the last step of this approximation, we take the $C_k$ to all be approximately equal in order to apply the geometric series. However, in reality we know $1/C_k$ to increase with $k$, giving large $k$ summands relatively more weight than in a geometric series. This would lead to a correction in the exponent.

\begin{equation}
    P_{corrected}(s)\approx A f(s)^{-1+\alpha},
\end{equation}
where $0<\alpha<<1$ \red{, I think this is because we want $\alpha$ to be a small correction, so in fact we'd want it to be much smaller than 1}.

\section*{Data and simulations}
 For large values of the chain length $N$ we thus find an approximate power law probability distribution for the sequence separation with an exponent of $-1+\alpha$. Here we use simulations, as introduced in \cite{molkenthin2020self} and networks extracted from measured protein data from the PDB \cite{PDB} to compare this behaviour in realistic 3D settings and show that the basic principle persists, indicating that the sequence distance separation distribution is largely determined by simple geometric constraints.

@@

\begin{figure}[h]
        \centering
	\includegraphics[width=0.45\textwidth]{figures/both_200.pdf}
	\includegraphics[width=0.45\textwidth]{figures/both_400.pdf}
        \caption{Sequence distance distributions for real and simulated folds for a) Chain lengths around 200 b) chain lengths around 400. The theoretical curve for 2D rings works okay for 3D open chains (because the rings are closed sequence separations of more than N/2 are a bit meaningless and thus not accounted for) The parameters used in this are $A=80$ and $\alpha=0.2$ in a) and $0.3$ in b). I thought in b) it makes sense for $\alpha$ to be larger as there are even larger exponents $k$ coming into relevance. I'm not sure we want to use that whole $\alpha$ argument too much though as it's a but ad-hoc and would be nicer to just fit $A$ and be done with it, but then it may look a bit more off. \red{Redo figure fancier, with more data, axis labels and possibly different lines (remove fit lines, maybe add just one 1/l heuristic line or something).}
        }
        \label{fig:sdd}
\end{figure}
We observe that, while the general shape seems to match, in real proteins sequence separations of 2 to 4 as well as 20 to 50 amino acids are slightly over-represented (see Fig.\ref{fig:sdd}) \red{I have a vague memory of figures with more data where the hump was less pronounced? Depending on this rephrase the above.}.

@@

\gray{We expect the mid-separation bump to be the result of secondary structure effects. While in the simulation all links are made with equal probability apart from geometric constraints, real protein chains have preferred bond angles leading to the energetic preference for $\alpha-$helix and $\beta-$sheet structures, within which connections only short connections can occur. Longer connections thus tend to be between different helices.
With helices ranging in length from a minimum of about 2 to 3 turns or 7-11 residues up to beyond 30 residues \cite{manning1988circular}, with the (rarer) longer helices accounting relatively for more connections, sequence separations of 20 to 50 seem likely to result from connections between subsequent helices. (I guess we could test this somehow, though i'm not sure how I would go about interpreting the results. I think the PDB files contain information on which residues belong to which structure, so we could do an analysis of how many connections of $s>4$ there are a) within one helix b) between a helix and the following loop) c) between 2 subsequent helices etc. Where I would count helices and sheets the same...)}
\begin{figure}[h]
        \centering
	\includegraphics[width=0.35\textwidth]{paper/figures/neighbouring_helices.pdf}
        \caption{\gray{The average length of a link between two subsequent helices is the average length of a helix plus the average length of a loop. \red{Does this suffice or do we want a fancy 3d image of this? If so, do you have the software that makes those?}
        \green{I can make a prettier thing provided we can actually make this argument based on the pdb data. I still struggle make the pdb data comply.}}
        }
        \label{fig:bumpexplain}
\end{figure}

\gray{For a more direct comparison between simulation results and PRNs, we have generated a sample of intrinsically disordered proteins \cite{something}, that also do not contain secondary structures. The results are shown in Fig.~\ref{fig:sdd_idp}.
In order to compare the 3D protein model simulations to PRNs, we selected protein structure lengths based on frequency and comparable length to the simulations. Fig.~\ref{fig:pdb_stats} shows that most proteins in the PDB have chain lengths between 100-400 protein residues. Therefore using example lengths of 200 and 400 residues seemed appropriate. From each of the over 20000 structures around the 200 and 400 residue length, we made the following choices for selecting different types of structures:
...}

\begin{figure}[h]
        \centering
	\includegraphics[width=0.45\textwidth]{figures/pdb_statistics.pdf}
        \caption{Statistics on all PDB entries based on the number of amino acid residues, with most entries found in the interval of 100-400 residues as of May 26th 2020. \red{should this figure be before the current fig 2? Because it's the reason we do 200 +- and 400 +-}
        }
        \label{fig:pdb_stats}
\end{figure}

\gray{We find that the over representation of connections with $20<s<50$ disappears, further confirming our expectation.}
\begin{figure}[h]
        \centering
	\includegraphics[width=0.45\textwidth]{figures/idp_200.pdf}
	\includegraphics[width=0.45\textwidth]{figures/idp_400.pdf}
        \caption{\gray{Sequence distance distributions for intrinsically disordered proteins and simulated folds for a) Chain lengths around 200 b) chain lengths around 400. The theoretical curve for 2D rings works okay for 3D open chains (because the rings are closed sequence separations of more than N/2 are a bit meaningless and thus not accounted for) The parameters used in this are $A=80$ and $\alpha=0.2$ in a) and $0.3$ in b). I thought in b) it makes sense for $\alpha$ to be larger as there are even larger exponents $k$ coming into relevance. I'm not sure we want to use that whole $\alpha$ argument too much though as it's a but ad-hoc and would be nicer to just fit $A$ and be done with it, but then it may look a bit more off.}
        }
        \label{fig:sdd_idp}
\end{figure}
\red{Anything else we want to say to that? This is a bit short now}
\section*{Conclusion}
We have demonstrated in analytical calculations and direct simulations how a random linking models with geometric constraints, such as the ones introduced in \cite{molkenthin2016scaling, molkenthin2020self} generate sequence distance distributions, that resemble a power-law, similar to the one found in folded proteins. This feature of protein network structure has been used to generate protein-like structures in \cite{bartoli2008effect}.

@@

\gray{For proteins with a high percentage of $\alpha$-helix structures, however, we find an over-representation of links with a sequence separation of 20 to 50 residues, implying a higher number of connections between neighbouring helices. We show that this is indeed the case by analyzing the sequence separation distributions of intrinsically disordered proteins. As helix structures arise from local bond angle preferences \cite{Danielsson2010}, rather than geometric constraints, these are not replicated in the model and thus the bump is absent in our theoretical calculations as well as simulations.}

\gray{This makes the sequence separation distribution an indicator for the disorderedness of a protein structure. If such a thing is interesting we could try to formalize this measure and see how good it is at identifying idp's, but i'm guessing they are quite easy to identify already.}

% \begin{enumerate}
%     \item Bartoli et al. made a random network model based on a power-law sequence separation distribution \cite{bartoli2008effect}
%     \item Here we show how such a distribution can be explained directly from geometric constraints
%     \item In the analysis of realistic sequence separation distributions from PRN's we find that the secondary structure leads to an over representation of $20<s<50$, which we can explain as connections between subsequent helices.
%     \item The bump is absent for idp's
%     \item This makes the sequence separation distribution an indicator for the disorderedness of a protein structure. If such a thing is interesting we could try to formalize this measure and see how good it is at identifying idp's, but i'm guessing they are quite easy to identify already.
% \end{enumerate}
\section*{Declarations}
\subsection{Availability of data and materials}
All data generated or analysed during this study are included in this published article [and its supplementary information files].
\subsection{Competing interests}
The authors declare no competing interests.
\subsection{Funding}
This research was supported by ...
\subsection{Authors' contributions}

\subsection{Acknowledgements}
We thank ... for fruitful discussions.

>>>>

%\documentclass{article}
%\usepackage[a4paper, total={6in, 8in}]{geometry}
%\documentclass[reprint,amsmath,amssymb,rmp,onecolumn,notitlepage,11pt]{revtex4-1}
\documentclass[
reprint,
twocolumn,
%preprint
%onecolumn,
amsmath,amssymb,superscriptaddress,aps,
pre]{revtex4-1}
\usepackage[utf8]{inputenc}
%\usepackage{authblk}
%\usepackage{natbib}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{xcolor}
\newcommand{\red}[1]{\textcolor{red!80!black}{#1}}
\newcommand{\blue}[1]{\textcolor{blue!80!black}{#1}}
\newcommand{\green}[1]{\textcolor{green!70!black}{#1}}
\newcommand{\gray}[1]{\textcolor{gray!80!black}{#1}}
\usepackage{mathtools}
\DeclarePairedDelimiter{\evdel}{\langle}{\rangle}
\usepackage[colorinlistoftodos]{todonotes}

\begin{document}
\title{Chain separation distribution in protein contact networks}
\title{Theoretical approximation of chain separation distribution in protein contact networks}
\title{Theory of chain separation distributions gives insights in real protein contact networks}
%\title{}
\author{Nora Molkenthin}
\affiliation{Potsdam Institut für Klimafolgenforschung}
\affiliation{Center for Advancing Electronics Dresden (cfaed), Technical University of Dresden}
\author{Steffen Mühle}
\affiliation{Uni Göttingen, 37077 Göttingen, Germany}
\author{Antonia S J S Mey}
\email[Electronic Address: ]{antonia.mey@ed.ac.uk}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}
%\author{Marc Timme}
% \affiliation{Network Dynamics, Max Planck Institute for Dynamics and Self-Organization (MPIDS), 37077 Göttingen, Germany}
% \affiliation{Chair for Network Dynamics, Institute for Theoretical Physics and Center for Advancing Electronics Dresden (cfaed), Technical University of Dresden, 01069 Dresden}

@@

Based on such considerations, the network formation process, repeatedly \red{(this may not be the adverb i'm looking for, but subsequently also doesnt quite work)} connects random pairs of units, without causing any intersections along the chain ~\cite{molkenthin2016scaling}. Such a model can now be used to investigate the emergent behaviour of sequence separation distributions as observed by Bartoli et al.~\cite{bartoli2008effect}. To this end, we analyze the sequence separation distributions that emerge from the geometric model, where the \emph{sequence separation} $\ell$ of a link is defined as the distance of the two end units along the original chain.
This is then compared to the sequence separation distribution extracted from protein contact networks generated from protein data bank data.

@@

\section*{Protein Residue Networks and their synthetic counterparts}
The complex interaction pattern between the amino acids in a protein can be very naturally expressed as a network, in which each amino acid is represented by a node and spatial proximity is encoded as a link. Whenever two central $C_\alpha$ atoms are closer together than a threshold $d_c$, they are connected. These connections are encoded in the adjacency matrix of the folded protein, which is the binary matrix
\begin{equation}
  A^{\textsf{PDB}}_{ij}=
  \begin{cases}
   0, & \text{ if } d_{i,j}>d_c \text{ or } i=j\\
      1, & \text{ if } d_{i,j}\leq d_c .
      \end{cases}
    \label{eq:aij}
\end{equation}
Adjacency matrices are commonly used to describe networks and in this case represent the PCN.

The underlying formation mechanisms of individual protein folds are incredibly complex and depend on the surrounding solvent as well as the specific amino acid sequence and their quantum-mechanical interactions. Here we want to study the ensemble of all proteins, however, and will thus "average" over such specifics.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/Fig1.pdf}
    \caption{a) The sequence separation distribution is defined as the separation along the chain (blue) of two connected units (red). b) The solid blue link prohibits some connections from node $A$ (dotted red), while others are still permitted (dotted blue). Counting all newly excluded links, we find that a link of length $k$ prohibits $2(l-1)$ links of length $l$ up to a maximum of $k$ and $2(k-1)$ links of length $k$ or more.}
    \label{fig:schematic}
\end{figure}

@@

In subsequent steps we get the same reduction but starting from the pool $F_{k-1}(s)$ of the step before rather than the full $N$ links for each length. The reduction also needs to be multiplied by a factor 
\begin{equation}
    P_k(s)=\frac{F_k(s)}{C_k},
    \label{eq.Pk}
\end{equation}
with $C_k=\sum_{s=2}^{N/2}F^k(s)$, to only remove links that are still in the pool. This leads to the recursion
\begin{align}
    F_k(s)&= \frac{1}{N/2-2} \nonumber \\
    &\sum_{s_k=2}^{N/2} {\begin{cases}
     F_{k-1}(s)-2(s-1) P_{k-1}(s) \text{ , for } s<s_k\\
     F_{k-1}(s)-2(s_k -1)P_{k-1}(s)\text{ , for } s\geq s_k
    \end{cases}}
\end{align}
which can be simplified to
\begin{align}
   F_k(s)&= \frac{1}{\frac{N}{2}-2} \frac{F_{k-1}(s)}{C_{k-1}}\nonumber \\
   &\left( \sum_{s_k=2}^{N/2}C_{k-1} - \sum_{s_k=2}^{s} 2(s_k-1) - \sum_{s_k=s+1}^{N/2} 2(s -1) \right) \nonumber \\
   &= \frac{1}{\frac{N}{2}-2}\frac{F_{k-1}(s)}{C_{k-1}}\nonumber \\
   &\left(\left(\frac{N}{2}-2\right)C_{k-1} -(s^2-s)
   -2\left(\frac{N}{2}-(s+1)\right)(s-1)\right)  \nonumber \\
   &= F_{k-1}(s)\left(1-\frac{-s^2 +(N-1)s-(N+1)}{(\frac{N}{2}-2)C_{k-1}} \right)\nonumber \\
   &=F_{k-1}(s)\left(1-\frac{f(s)}{(\frac{N}{2}-2)C_{k-1}} \right)
   \label{eq.Fk_rec}
\end{align}
Where we defined $f(s)=-s^2 +(N-1)s-(N+1) \approx N s - s^2 - N$ for notational brevity.

@@

In the last step of this approximation, we take the $C_k$ to all be approximately equal in order to apply the geometric series. However, in reality we know $1/C_k$ to increase with $k$, giving large $k$ summands relatively more weight than in a geometric series. This can be expressed as a small correction $\alpha$ in the exponent.

\begin{equation}
    P_{corrected}(s)\approx A f(s)^{-1+\alpha},
\end{equation}
where $0<\alpha \ll 1$.

\section*{Data and simulations}
 For large values of the chain length $N$ we thus find an approximate power law probability distribution for the sequence separation with an exponent of $-1+\alpha$. Here we use simulations, as introduced in \cite{molkenthin2020self} and networks extracted from measured protein data from the PDB \cite{PDB} to compare this behaviour in realistic 3D settings and show that the basic principle persists, indicating that much of the sequence distance separation distribution can be explained by simple geometric constraints.
 \begin{figure}[h]
        \centering
	\includegraphics[width=0.45\textwidth]{figures/pdb_statistics.pdf}
        \caption{Statistics on all PDB entries based on the number of amino acid residues, with most entries found in the interval of 100-400 residues as of May 26th 2020.}
        \label{fig:pdb_stats}
\end{figure}

@@

\begin{figure*}[t]
        \centering
	\includegraphics[width=\textwidth]{paper/figures/resultsfig_old.pdf}
        \caption{Sequence distance distributions for real and simulated folds for a) Chain lengths around 200 b) chain lengths around 300 c) chain lengths around 400.
        }
        \label{fig:sdd}
\end{figure*}
We observe that, while the general shape seems to match, in real proteins sequence separations of 2 to 4 as well as 20 to 50 amino acids are slightly over-represented (see Fig.\ref{fig:sdd}) \red{I have a vague memory of figures with more data where the hump was less pronounced? Depending on this rephrase the above.}.

@@

\section*{Conclusion}
We have demonstrated in analytical calculations and direct simulations how a random linking models with geometric constraints, such as the ones introduced in \cite{molkenthin2016scaling, molkenthin2020self} generate sequence distance distributions, that resemble a power-law, similar to the one found in folded proteins. This feature of protein network structure has been used to generate protein-like structures in \cite{bartoli2008effect}.

@@

\section*{Declarations}
\subsection{Availability of data and materials}
All data generated or analysed during this study are included in this published article [and its supplementary information files].
\subsection{Competing interests}
The authors declare no competing interests.
\subsection{Funding}
This research was supported by ...
\subsection{Authors' contributions}

\subsection{Acknowledgements}
We thank Marc Timme for fruitful discussions.
